---
title: "PSL Project 4: Movie Recommendation"
author: "Vijayakumar Sitha Mohan (VS24), Waitong Matthew Leung (wmleung2)"
date: '30-Nov-2020'
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Introduction

We are provided with a dataset contains about 1 million anonymous ratings of approximately 3,900 movies made by 6,040 MovieLens users who joined MovieLens in 2000. The goal is to use the rating data to build a movie recommendation system based on a few different recommendation schemes, namely System I has two schemes based on movie genres, and System II has two schemes based on collaborative recommendation schemes.


System I is based on movie genres. 

- One scheme recommend top 5 popular movies in each user selected genres. Popularity is measured by the number of ratings a movie receives. A popular movie may not be a highly rated movie. 

- The other scheme recommend 5 random movies in each user selected genres.



System II is based on collaborative filtering recommendation. 

- One scheme uses user-based collaborative filtering technique to estimate missing movie ratings based on other similar users ratings. Once missing movie ratings are estimated, movies with highest ratings is recommended at the top. 

- The other scheme uses item-based collaborative filtering technique to estimate missing movie ratings based on other movies similar to the onces that rated highly by the user. The most similar one is recommended to the top.


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
library(recommenderlab)
library(Matrix)
library(tidytable)
library(knitr)
```

## Download Data

Download the MovieLens 1M Dataset. There are four files: 

 * `README` [[Github_link](https://liangfgithub.github.io/MovieData/README?raw=true)]
 * `movies.dat` [[Github_link](https://liangfgithub.github.io/MovieData/movies.dat?raw=true)]
 * `ratings.dat` [[Github_link](https://liangfgithub.github.io/MovieData/ratings.dat?raw=true)]
 * `users.dat` [[Github_link](https://liangfgithub.github.io/MovieData/users.dat?raw=true)]
 
 
```{r}
 myurl = "https://liangfgithub.github.io/MovieData/"
```
 

Check the readme file to understand the format of the other three files. You can also download them from the original site: [https://grouplens.org/datasets/movielens/](https://grouplens.org/datasets/movielens/).



## Read in Data

### ratings data
```{r}
# use colClasses = 'NULL' to skip columns
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')

# TO DO: comment out the following line before submit. It is to limit the number of ratings item to speed up testing
#ratings = ratings[1:100000,]
```

```{r}
write.csv(ratings,file="ratings.csv",quote=FALSE,row.names = FALSE)
```

```{r, result='asis', echo=FALSE}
datatable(head(ratings, 6), 
          class = "nowrap hover row-border", 
          options = list(dom = 't',
                         scrollX = FALSE, 
                         autoWidth = TRUE))
```

### movies data
In movies.dat, some movie names contain single colon (:), so the method above does not work. 

```{r}
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))

```
```{r}
write.csv(movies,file="movies.csv",quote=FALSE)
```

```{r}
# movies = movies['Genres'].str.get_dummies('|')
```


```{r, result='asis', echo=FALSE}
datatable(head(movies, 6), 
          class = "nowrap hover row-border", 
          options = list(dom = 't',
                         scrollX = FALSE, 
                         autoWidth = TRUE))
```

### user data
```{r}
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
```

```{r, result='asis', echo=FALSE}
datatable(head(users, 6), 
          class = "nowrap hover row-border", 
          options = list(dom = 't',
                         scrollX = FALSE, 
                         autoWidth = TRUE))
```

For users, `Gender` is denoted by "M" for male and "F" for female, `Age` is chosen from the following ranges:

	*  1:  "Under 18"
	* 18:  "18-24"
	* 25:  "25-34"
	* 35:  "35-44"
	* 45:  "45-49"
	* 50:  "50-55"
	* 56:  "56+"

and `Occupation` is chosen from the following choices:

	*  0:  "other" or not specified
	*  1:  "academic/educator"
	*  2:  "artist"
	*  3:  "clerical/admin"
	*  4:  "college/grad student"
	*  5:  "customer service"
	*  6:  "doctor/health care"
	*  7:  "executive/managerial"
	*  8:  "farmer"
	*  9:  "homemaker"
	* 10:  "K-12 student"
	* 11:  "lawyer"
	* 12:  "programmer"
	* 13:  "retired"
	* 14:  "sales/marketing"
	* 15:  "scientist"
	* 16:  "self-employed"
	* 17:  "technician/engineer"
	* 18:  "tradesman/craftsman"
	* 19:  "unemployed"
	* 20:  "writer"

## Exploration

### Unique Users? 
UserID ranges from 1 to 6040. 
```{r}
dim(users)
length(unique(ratings$UserID))
```

### Unique Movies? 
MovieID ranges from 1 to 3952, but not all of them are included in `movies`. Further,  177 movies are in `movies` but not rated in `ratings`. 
```{r}
dim(movies)
length(unique(ratings$MovieID))
movies_not_rated = movies %>% 
  filter(!(MovieID %in% ratings$MovieID))
dim(movies_not_rated)
```

### Dist of Ratings

Most of the ratings are in the 3-5 range. 
```{r}
tmp = data.frame(Rating = 1:5, 
                 freq = as.vector(table(ratings$Rating)/nrow(ratings)))
ggplot(data = tmp, aes(x = Rating, y = freq)) +
  geom_bar(stat="identity", fill = 'steelblue', width = 0.6) + 
  geom_text(aes(label=round(freq, dig=2)), 
                vjust=1.6, color="white", size=3.5) +
  theme_minimal()
```

### Ratings per User

Most users have left less than 500 ratings; about 9 users have had over 1300 ratings; one has 2314 ratings. 

```{r eval=TRUE}
tmp = ratings %>% 
  group_by(UserID) %>% 
  summarize(ratings_per_user = n()) 
summary(tmp$ratings_per_user)
stem(tmp$ratings_per_user)
sum(tmp$ratings_per_user > 500)
sort(tmp$ratings_per_user[tmp$ratings_per_user>1300])
```

```{r}
tmp %>%
  ggplot(aes(ratings_per_user)) +
  geom_bar(fill = "steelblue") + coord_cartesian(c(20, 500))
```

Combining `users` and `tmp`, you could further explore how `ratings_per_user` depends on Gender, Age, and Occupation of users. 

```{r message=FALSE, warning=FALSE}
tmp = tmp %>% full_join(users, by = 'UserID')
```

### Ratings per Movie

There are 31 movies that have received more than 2000 ratings. The most popular movie is "American Beauty (1999)", followed by the "Star Wars" series. Throughout, **popular** means receiving many ratings; a popular movie may not be a **highly-rated** movie. 

```{r}
tmp = ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), ave_ratings = mean(Rating)) %>%
  inner_join(movies, by = 'MovieID')
summary(tmp$ratings_per_movie)
```

```{r}
tmp %>% 
  filter(ratings_per_movie > 2000) %>%
  arrange(desc = ratings_per_movie) %>%
  select(c("Title", "ratings_per_movie")) %>%
  print(n = 31)

tmp %>% ggplot(aes(ratings_per_movie)) + 
  geom_bar(fill = "steelblue", width = 1) + coord_cartesian(c(1,1500))
```

The top ten highly-rated (based on their average ratings) among all movies that have received at least 1000 ratings.

```{r}
small_image_url = "https://liangfgithub.github.io/MovieImages/"
ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), 
            ave_ratings = round(mean(Rating), dig=3)) %>%
  inner_join(movies, by = 'MovieID') %>%
  filter(ratings_per_movie > 1000) %>%
  top_n(10, ave_ratings) %>%
  mutate(Image = paste0('<img src="', 
                        small_image_url, 
                        MovieID, 
                        '.jpg?raw=true"></img>')) %>%
  select('Image', 'Title', 'ave_ratings') %>%
  arrange(desc(-ave_ratings)) %>%
  datatable(class = "nowrap hover row-border", 
            escape = FALSE, 
            options = list(dom = 't',
                          scrollX = TRUE, autoWidth = TRUE))
```

### Dist of Genres

First create a binary indicator for the 18 genres for each movie.  
```{r message=FALSE, warning=FALSE}
genres = as.data.frame(movies$Genres, stringsAsFactors=FALSE)
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
genre_list = c("Action", "Adventure", "Animation", 
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical", 
               "Mystery", "Romance", "Sci-Fi", 
               "Thriller", "War", "Western")
m = length(genre_list)
genre_matrix = matrix(0, nrow(movies), length(genre_list))
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
}
colnames(genre_matrix) = genre_list
remove("tmp", "genres")
```

Then we can output histograms of the 18 genres based on `movies` or based on `ratings`.

```{r}
data.frame(Genres = genre_list, 
                 Freq = as.vector(colMeans(genre_matrix))) %>% 
  ggplot(aes(reorder(Genres, Freq), Freq, fill = Freq)) + 
  geom_bar(stat = "identity") + 
    geom_text(aes(label = round(Freq, dig=2)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  scale_colour_brewer(palette="Set1") + 
  labs(y = 'Frequency', x = 'Genre')
```

```{r}
tmp = ratings %>% 
  left_join(data.frame(MovieID = movies$MovieID, genre_matrix), 
            by = "MovieID") %>%
  select(-c("UserID", "MovieID", "Rating", "Timestamp"))

data.frame(Genres = genre_list, 
                 Popularity = as.vector(colMeans(tmp))) %>% 
  ggplot(aes(reorder(Genres, Popularity), Popularity, fill = Popularity)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(Popularity, dig=3)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  labs(y = 'Popularity', x = 'Genre')
```


About half of the movies belong to only one genre; about one-third are categorized into two genres; a few are categorized into more than four genres. The movie "The Transformers" (1986) are categorized into six genres: Action, Animation, Children's, Sci-Fi, Thriller, and War. 

```{r}
tmp = colSums(genre_matrix)
summary(tmp)
movies[which(tmp==6), ]
movies[which(tmp==5), ]
```

For illustration purpose only, let's assume `movies` contains all the movies available to users from 1919 to 2000. Then we can compute the cumulative percentages of the 18 genres over year from 1919 to 2000 and store them in the 81-by-19 matrix `tmp`. For example, till 2000, users can access about 7.8\% Action, 4.4\% Adventure, 25\% Drama, etc. A graphical display of such CDF over 10 generes are displayed below.  
```{r}
# range(movies$Year) % 1919 to 2000
tmp = data.frame(Year = movies$Year, genre_matrix) %>%
  group_by(Year) %>%
  summarise_all(sum)
tmp[,-1] = apply(tmp[, -1], 2, cumsum)
tmp[,-1] = tmp[,-1]/sum(tmp[nrow(tmp), -1])
print(round(tmp[nrow(tmp),-1], dig=3))
```



```{r message=FALSE, warning=FALSE}
tmp = reshape2::melt(tmp, id.vars="Year") 
tmp %>%
  ggplot(aes(Year, value, group = variable)) +
  geom_area(aes(fill = variable)) + 
  geom_line(aes(group = variable), position = "stack")
```  



```{r}
ratings$Timestamp = NULL
```

## Train and Test Split

Create

- train data that contain about 80% rows of ratings.dat

- test data that contain about 20% of rows of ratings.dat



```{r}
set.seed(100)
train.id = sample(nrow(ratings), floor(nrow(ratings)) * 0.8)
train = ratings[train.id, ]
head(train)

test = ratings[-train.id, ]
head(test)
```

## Recommender System


### First create a Utility matrix stroed as a Sparse matrix

```{r}
i = paste0('u', train$UserID)
j = paste0('m', train$MovieID)
x = train$Rating
tmp = data.frame(i, j, x, stringsAsFactors = T)
Rmat = sparseMatrix(as.integer(tmp$i), as.integer(tmp$j), x = tmp$x)
rownames(Rmat) = levels(tmp$i)
colnames(Rmat) = levels(tmp$j)
Rmat = new('realRatingMatrix', data = Rmat)
```


Rmat is a 6040-by-3681 sparse matrix. Its rows correspond to the unique 6040 users in the training data, and columns correspond to the unique 3681 movies in the training data (although the MovieIDs range from 1 to 3952).



```{r}
range(train$MovieID)
length(unique(train$MovieID))
range(train$UserID)
length(unique(train$UserID))
dim(Rmat)
```




### System Ia - Genre based (Average Popular Rating)


- Train a genre based recommender system by popularity using R package [recommenderlab]. Popularity is defined as higher number of reviews rather then highly rated reviews.

```{r}
rec_POPULAR = Recommender(Rmat, method = 'POPULAR',
                          parameter = list(normalize = 'center'))

#Summary of model parameters
rec_POPULAR@model
```

- Make prediction on the test data. We are get the predicted ratings from the model first. We average the predicted ratings for every single movies. We ignore any NA return from the model. We use this average popular ratings to assign to recommend to every single users.


```{r}
# predict ratings for all users, and compare it to first three users
start.time = Sys.time()
POPULAR_recom = predict(rec_POPULAR, 
                Rmat, type = 'ratingMatrix')
recom_mat <- as(Rmat, "matrix")
AVG_POPULAR_recom = colMeans(recom_mat,na.rm=TRUE)
POPULAR_time = Sys.time() - start.time

```

We display the predicted average popular ratings for every users, and original rating of the first 10 movies for the three users.


```{r}

AVG_POPULAR_recom[1:10]
as(Rmat, 'matrix')[1:3, 1:10]
```



- Evaluation with RMSE

We replace any NA predicted rating with weighted average of original user rating distribution of 3.61.

```{r, eval=TRUE}
# This may take a long time
#start.time = Sys.time()
#POPULAR_recom = predict(rec_POPULAR, 
#                Rmat, type = 'ratings')  
#POPULAR_rec_list = as(POPULAR_recom, 'list')  # each element are ratings of that user

POPULAR_test.pred = test
POPULAR_test.pred$rating = NA

#POPULAR_time = Sys.time() - start.time
```

```{r}

# For all lines in test file, one by one
for (u in 1:nrow(test)){

    # Read userid and movieid from columns 2 and 3 of test data
    movieid = paste("m",as.character(test$MovieID[u]), sep ="" )

    # handle missing values; replace with weighted average of original user rating distribution of 3.61
    POPULAR_test.pred$rating[u] = ifelse(movieid %in% names(AVG_POPULAR_recom), AVG_POPULAR_recom[[movieid]], 3.61) 
}

# Calculate RMSE
POPULAR_RMSE = sqrt(mean((test$Rating - POPULAR_test.pred$rating)^2)) 

```


### System Ib - Genre based (Hybrid)


- Train a genre based recommender system with hybrid model using R package [recommenderlab]. The model mix with POPULAR movie (most reviewed), RANDOM for serendipity and RERECOMMEND for the user his/her own rating with weighting of 0.6, 0.2 and 0.2 respectively.

```{r}
# mix popular movies with a random recommendations for diversity and rerecommend some movies the user liked.
rec_HYBRID = HybridRecommender(
  Recommender(Rmat, method = "POPULAR"),
  Recommender(Rmat, method = "RANDOM"),
  Recommender(Rmat, method = "RERECOMMEND"),
  weights = c(.6, .2, .2)
)


#Summary of model parameters
rec_HYBRID@model
```

- Make prediction on the test data

First let's predict all un-rated movies for the first three users in the training data. No information for users not in the training data; the coldstart issue. 

```{r}
# predict ratings for the first three users
HYBRID_recom = predict(rec_HYBRID, 
                Rmat[1:3], type = 'ratings')
```

We display the original and predicted ratings of the first 10 movies for the three users. Note that the final prediction could still have missing values. 

```{r}
as(HYBRID_recom, 'matrix')[, 1:10]
as(Rmat, 'matrix')[1:3, 1:10]
```



- Evaluation with RMSE

We replace any NA predicted rating with weighted average of original user rating distribution of 3.61.

```{r, eval=TRUE}
# This may take a long time
start.time = Sys.time()
HYBRID_recom = predict(rec_HYBRID, 
                Rmat, type = 'ratings')  
HYBRID_rec_list = as(HYBRID_recom, 'list')  # each element are ratings of that user

HYBRID_test.pred = test
HYBRID_test.pred$rating = NA

HYBRID_time = Sys.time() - start.time
```

```{r}

# For all lines in test file, one by one
for (u in 1:nrow(test)){

    # Read userid and movieid from columns 2 and 3 of test data
    userid = as.integer(test$UserID[u])
    movieid = as.integer(test$MovieID[u])
    
    rating = HYBRID_rec_list[[userid]][movieid]
    # handle missing values; replace with weighted average of original user rating distribution of 3.61
    HYBRID_test.pred$rating[u] = ifelse(is.na(rating), 3.61, rating)
}

# Calculate RMSE
HYBRID_RMSE = sqrt(mean((test$Rating - HYBRID_test.pred$rating)^2)) 
HYBRID_RMSE
```





### System IIa - User-Based Collaborative Filtering


Train a user-based collaborative filtering (UBCF) recommender system using R package [recommenderlab]. 

We use the following parameters for the UBCF model:

  - Z-score normalization to the ratings matrix
  
  - pick 5 nearest neighbor
  
  - similarity measure is cosine similarity

```{r}
rec_UBCF = Recommender(Rmat, method = 'UBCF',
                  parameter = list(normalize = 'Z-score', 
                                   method = 'Cosine', 
                                   nn = 25))
#Summary of model parameters
rec_UBCF@model
```

- Make prediction on the test data

First let's predict all un-rated movies for the first three users in the training data. No information for users not in the training data; the coldstart issue. 

```{r}
# predict ratings for the first three users
UBCF_recom = predict(rec_UBCF, 
                Rmat[1:3], type = 'ratings')
```

We display the original and predicted ratings of the first 10 movies for the three users. Note that the final prediction could still have missing values. 

```{r}
as(UBCF_recom, 'matrix')[, 1:10]
as(Rmat, 'matrix')[1:3, 1:10]
```

- Evaluation with RMSE

We replace any NA predicted rating with weighted average of original user rating distribution of 3.61.

```{r, eval=TRUE}
# This may take a long time
start.time = Sys.time()
UBCF_recom = predict(rec_UBCF, 
                Rmat, type = 'ratings')  
UBCF_rec_list = as(UBCF_recom, 'list')  # each element are ratings of that user

UBCF_test.pred = test
UBCF_test.pred$rating = NA

UBCF_time = Sys.time() - start.time
```

```{r}

# For all lines in test file, one by one
for (u in 1:nrow(test)){

    # Read userid and movieid from columns 2 and 3 of test data
    userid = as.integer(test$UserID[u])
    movieid = as.integer(test$MovieID[u])
    
    rating = UBCF_rec_list[[userid]][movieid]
    # handle missing values; replace with weighted average of original user rating distribution of 3.61
    UBCF_test.pred$rating[u] = ifelse(is.na(rating), 3.61, rating)
}

# Calculate RMSE
UBCF_RMSE = sqrt(mean((test$Rating - UBCF_test.pred$rating)^2)) 
UBCF_RMSE
```



```{r}
# recommenderRegistry$get_entry_names()
```


### System IIb - Item-based Collaborative Filtering

Train Item-based collaborative filtering (IBCF) recommender system using R package recommenderLab

We use the following parameters for the IBCF model:

  - Z-score normalization to the ratings matrix
  
  - similarity measure is cosine similarity
  
  - similarity matrix is not normalized

```{r message=FALSE, warning=FALSE}
rec_IBCF = Recommender(Rmat, method = 'IBCF',
                  parameter = list(normalize = 'Z-score', 
                                   method = 'Cosine'))


#Summary of model parameters
rec_IBCF@model$method
rec_IBCF@model$normalize
rec_IBCF@model$normalize_sim_matrix

```

```{r}
model_details <- getModel(rec_IBCF)
model_details$description
n_items_top = 20
#image(model_details$sim[1:n_items_top, 1:n_items_top], main = "Heatmap of the first rows and columns")

```


- Make prediction on the test data

First we predict the top 5 recommended movie for the first user.


```{r}
recc_predicted <- predict(object = rec_IBCF, newdata = Rmat[1:100], n = 5)
recc_predicted
class(recc_predicted)
slotNames(recc_predicted)
recc_predicted@items[[1]]
```

This is the predicted titles.


```{r}
recc_user_1 <- recc_predicted@items[[1]]
movies_user_1 <- recc_predicted@itemLabels[recc_user_1]
movies[movies$MovieID %in% recc_user_1,]$Title
```

```{r message=FALSE, warning=FALSE}
recc_matrix <- sapply(recc_predicted@items, function(x){
  movies[movies$MovieID == x,]

  
})
dim(recc_matrix)
recc_matrix[, 1:4]
```

- Evaluation with RMSE

We replace any NA predicted rating with weighted average of original user rating distribution of 3.61.

```{r, eval=TRUE}
# This may take a long time
start.time = Sys.time()
IBCF_recom = predict(rec_IBCF, 
                Rmat, type = 'ratings')  
IBCF_rec_list = as(IBCF_recom, 'list')  # each element are ratings of that user

IBCF_test.pred = test
IBCF_test.pred$rating = NA

IBCF_time = Sys.time() - start.time
```

```{r}

# For all lines in test file, one by one
for (u in 1:nrow(test)){

    # Read userid and movieid from columns 2 and 3 of test data
    userid = as.integer(test$UserID[u])
    movieid = as.integer(test$MovieID[u])
    
    rating = IBCF_rec_list[[userid]][movieid]
    # handle missing values; replace with weighted average of original user rating distribution of 3.61
    IBCF_test.pred$rating[u] = ifelse(is.na(rating), 3.61, rating)
}

# Calculate RMSE
IBCF_RMSE = sqrt(mean((test$Rating - IBCF_test.pred$rating)^2))
IBCF_RMSE
```

## Summary

From the following table of prediction speed and lowest RMSE, System I Average Popular model stands out.


RMSE Comparison of different System and scheme

```{r echo=FALSE}
results = data.frame("System I: Avg Popular" = POPULAR_RMSE, "System I: Hybrid" = HYBRID_RMSE, "System II: UBCF"= UBCF_RMSE, "System II: IBCF"= IBCF_RMSE)

kable(results)
```

Prediction running time Comparison of different System and scheme

```{r echo=FALSE}

results2 = data.frame("System I: Avg Popular" = POPULAR_time, "System I: Hybrid" = HYBRID_time, "System II: UBCF"= UBCF_time, "System II: IBCF"= IBCF_time)

kable(results2)

```

