---
title: "PSL (F20) Coding Assignment 3"
author: "Vijayakumar Sitha Mohan (VS24), Waitong Matthew Leung (wmleung2)"
date: '12-Oct-2020'
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
set.seed(8742)
```

## Introduction

Write my own functions to use LOO-CV and GCV to select the optimal span for loess.

- Test my code on data set Coding3 Data.csv, which we downloaded from Piazza.

- Report my CV and GCV for 15 span values: 0.20; 0.25;... ; 0.90.

- Report the optimal value(s) for span based on CV and GCV.

- Plot the fitted curve(s) using the optimal value(s) for span.


## Prepare my function

Write my own function to use LOO-CV and GCV to select the optimal span for loess.

```{r}

lo.lev <- function(x1, sp){
  # x1: feature vector of length n
  # sp: a value for "span"
  
  n = length(x1);
  lev = rep(0, n)
  
  ##############################################
  # YOUR CODE: Compute the diagonal entries of the 
  #            smoother matrix S and 
  #            store it in a vector "lev"
  # Tip: check how we compute the smoother matrix
  #      for smoothing spline models
  
  for(i in 1:n){
    y = rep(0, n); y[i]=1;
    yi = loess(y ~ x1, span = sp)$fitted;
    lev[i]= yi[i];
  }
  
  ##############################################
  
  
  return(lev)
}

onestep_CV <- function(x1, y1, sp){
  
  ##############################################
  #  YOUR CODE: 
  #  1) Fit a loess model y1 ~ x1 with span = sp, and extract 
  #     the corresponding residual vector
  #  2) Call lo.lev to obtain the diagonal entries of S
  #  3) Compute LOO-CV and GCV using formula from lecture notes
  #    [lec_W5_NonlinearRegression.pdf] page 33. 
  
  n = length(x1)
  err = y1 - loess(y1 ~ x1, span = sp)$fitted
  sdiag = lo.lev(x1, sp)
  
  cv = sum((err/(1-sdiag))^2)/n
  gcv = sum((err/(1-sum(sdiag)/n))^2)/n
  
  ##############################################
  
  return(list(cv = cv, gcv = gcv))
}

myCV <- function(x1, y1, span){
  # x1: feature vector of length n
  # y1: response vector of length n
  # span: a sequence of values for "span"
  
  m = length(span)
  cv = rep(0, m)
  gcv = rep(0, m)
  
  for(i in 1:m){
    tmp = onestep_CV(x1, y1, span[i])
    cv[i] = tmp$cv
    gcv[i] = tmp$gcv
  }
  return(list(cv = cv, gcv = gcv))
}

```



## Test my function

Test my function with data "Coding3_Data.csv."

```{r}

mydata = read.csv(file = "Coding3_Data.csv")
dim(mydata)


```


```{r}
plot(mydata$x, mydata$y, xlab="", ylab="")

```


Create a grid of values for span: 15 values that are equally spaced between 0.20 and 0.90. Call your function myCV to compute the corresponding LOO-CV and GCV.

```{r}

span1 = seq(from = 0.2, by = 0.05, length = 15 )
cv.out = myCV(mydata$x, mydata$y, span1)

```


## Results : Print out my results

Print my results on LOO-CV and GCV. Both achieve their minimal at 0.5.

```{r}

myout = data.frame(CV = cv.out$cv, 
                   GCV = cv.out$gcv, 
                   span = span1)
myout

```


```{r}

myout$span[myout$GCV == min(myout$GCV)]

```


```{r}

myout$span[myout$CV == min(myout$CV)]

```


## Plot the fitted curve

Plot the data (red circles), the true curve (gray) and the fitted curve (blue dashed line) using the optimal span.

```{r}

spangcv.min = 0.5
plot(mydata$x, mydata$y, xlab="", ylab="", col="red");
fx = 1:50/50;
fy = sin(12*(fx+0.2))/(fx+0.2)
lines(fx, fy, col=8, lwd=2, color = "gray");

f = loess(y ~ x, mydata, span = spangcv.min)
lines(fx, predict(f, newdata = data.frame(x = fx), surface = "direct"), lty=2, lwd=2, col="blue")

```


## Discussion

Both LOO-CV and GCV select the same optimal span for Loess, i.e. 0.5, for the dataset of this assignment. The fitted values with this optimal span looks pretty close to the real signal.


## Contribution

Vijayakumar Mohan contributes to verification of implementation and enhancement of visualization.

Matthew Leung contributes to implementation of LOO-CV/GCV and overall presentation of the report
