---
title: "PSL (F20) Project 2"
author: "Vijayakumar Sitha Mohan (VS24), Waitong Matthew Leung (wmleung2)"
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: no
urlcolor: cyan
fontsize: 10pt
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```


```{r echo = FALSE, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)
library(dplyr)
library(tidyr)
library(kableExtra)
```

#### Introduction

- The goal is to predict the future weekly sales for each department in each store based on the historical data for 45 Walmart Stores located in different regions. With given train_ini.csv, the data till 2011-02, we need to predict the weekly sales for 2011-03 and 2011-04. After that provided with the weekly sales data for 2011-03 and 2011-04 (fold_1.csv), and we need to predict the weekly sales for 2011-05 and 2011-06, and so on.

- t = 1, predict 2011-03 to 2011-04 based on data from 2010-02 to 2011-02 (train_ini.csv);

- t = 2, predict 2011-05 to 2011-06 based on data from 2010-02 to 2011-04 (train_ini.csv, fold_1.csv);

- t = 3, predict 2011-07 to 2011-08 based on data from 2010-02 to 2011-06 (train_ini.csv, fold_1.csv, fold_2.csv);

- ……

- t = 10, predict 2011-09 to 2011-08 baesd on data from 2010-02 to 2011-08 (train_ini.csv, fold_1.csv, fold_2.csv, …, fold_9.csv)

#### Overall Approach


- Our approach was an extension to the solution(III) provided by the professor which trains the linear model for each Store, Dept combination. We use the historical weekly sales data to predict future Weekly_Sales (response variable Y) by Store/Dept combination. 

- Since we train each linear model with only data from the same Store and Dept, we don't need to include Store/Dept in the predictor.

- We derive two predictors, namely "Yr" and "Wk", from original data column Date. "Wk" is the number of the week in each year starting from 1 to 52(or 53). "Yr" is the year from data column Date. We train a linear regression model: sqrt(Y) ~ Yr + Wk for each combination of Store/Dept. We apply square root transformation to response variable Y to reduce skewness of data. It inflates smaller Y but stabilizes bigger Y. 


#### Data PreProcessing


- The following preprocessing applied by the professor in the given solution(iii)

    + Make sure each year to have exactly 52 weeks

    + Not all stores and departments needed prediction

    + extract year to predictor "Yr" from Date

    + extract week to predictor "Wk" from Date

    + check if "Yr" == 2010, subtract 1 from "Wk" since there is one more week in 2010. This is to line up the week from last year. The most important thing is to line up holiday weekend as a result.

    + set "Wk" as factor variable with 52 levels, i.e. one per week.

    + since there is some missing data, we use model.matrix() to build design matrix and replace any NA with zero before feeding to lm() to prevent erroring out lm().

- In addition, we applied

    + Replace any negative response variable Weekly_Sales with zero.

    + Apply Square root tranformation for response variable Weekly Sales.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# read raw data and extract date column
train_raw <- readr::read_csv(unz('train.csv.zip', 'train.csv'))
train_dates <- train_raw$Date

# training data from 2010-02 to 2011-02
start_date <- ymd("2010-02-01")
end_date <- start_date %m+% months(13)

# split dataset into training / testing
train_ids <- which(train_dates >= start_date & train_dates < end_date)
train = train_raw[train_ids, ]
test = train_raw[-train_ids, ]

# create the initial training data
readr::write_csv(train, 'train_ini.csv')

# create test.csv
# removes weekly sales
test %>%
  select(-Weekly_Sales) %>%
  readr::write_csv('test.csv')

# create 10-fold time-series CV
num_folds <- 10
test_dates <- train_dates[-train_ids]

# month 1 --> 2011-03, and month 20 --> 2012-10.
# Fold 1 : month 1 & month 2, Fold 2 : month 3 & month 4 ...
for (i in 1:num_folds) {
    # filter fold for dates
    start_date <- ymd("2011-03-01") %m+% months(2 * (i - 1))
    end_date <- ymd("2011-05-01") %m+% months(2 * (i - 1))
    test_fold <- test %>%
        filter(Date >= start_date & Date < end_date)
   
    # write fold to a file
    readr::write_csv(test_fold, paste0('fold_', i, '.csv'))
}

```

```{r echo = FALSE, message=FALSE, warning=FALSE}
t = 1
train <- readr::read_csv('train_ini.csv')
test <- readr::read_csv('test.csv')
```

- Scatter plot to show Outlier & Skewness in Weekly Sales of training dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow = c(1,2))
hist(train$Weekly_Sales, col = 'light green', main = "Weekly Sales Original", xlab = "Weekly Sales", breaks=50)
hist(sqrt(train$Weekly_Sales), col = 'light green', main = "Weekly Sales Transformed", xlab = "Weekly Sales",breaks=50)
```


```{r echo = FALSE, message=FALSE, warning=FALSE}
mypredict = function(){
  if (t>1){
    train <<- train %>% add_row(new_train)
  }
 
  start_date <- ymd("2011-03-01") %m+% months(2 * (t - 1))
  end_date <- ymd("2011-05-01") %m+% months(2 * (t - 1))
 
  test_current <- test %>%
    filter(Date >= start_date & Date < end_date) %>%
    select(-IsHoliday)
  
  # Convert negative Weekly Sales to 0
  train$Weekly_Sales[train$Weekly_Sales < 0] = 0
  #train$Weekly_Sales[train$Weekly_Sales == 0] = 1
 

# not all depts need prediction
  test_depts <- unique(test_current$Dept)
  test_pred <- NULL
 
  for(dept in test_depts){
    train_dept_data <- train %>% filter(Dept == dept)
    test_dept_data <- test_current %>% filter(Dept == dept)
   
   # no need to consider stores that do not need prediction
   # or do not have training samples
    train_stores <- unique(train_dept_data$Store)
    test_stores <- unique(test_dept_data$Store)
    test_stores <- intersect(train_stores, test_stores)
   
    for(store in test_stores){
      tmp_train <- train_dept_data %>%
        filter(Store == store) %>%
        mutate(Wk = ifelse(year(Date) == 2010, week(Date)-1, week(Date))) %>%
        mutate(Yr = year(Date))
      tmp_test <- test_dept_data %>%
        filter(Store == store) %>%
        mutate(Wk = ifelse(year(Date) == 2010, week(Date)-1, week(Date))) %>%
        mutate(Yr = year(Date))

      tmp_train$Wk = factor(tmp_train$Wk, levels = 1:52)
      tmp_test$Wk = factor(tmp_test$Wk, levels = 1:52)
      train_model_matrix <- model.matrix(~ Yr + Wk, tmp_train)
      test_model_matrix <- model.matrix(~ Yr + Wk, tmp_test)
     

   
      mycoef <- lm(sqrt(tmp_train$Weekly_Sales) ~ train_model_matrix)$coef
      mycoef[is.na(mycoef)] <- 0
      tmp_pred <- mycoef[1] + test_model_matrix %*% mycoef[-1]
      tmp_pred = tmp_pred^2
 
      tmp_test <- tmp_test %>%
        mutate(Weekly_Pred = tmp_pred[,1]) %>%
        select(-Wk, -Yr)
      test_pred <- test_pred %>% bind_rows(tmp_test)
   }
  }
  return(test_pred)
}
```



```{r echo=FALSE, message=FALSE, warning=FALSE}
# save weighted mean absolute error WMAE
start.time = Sys.time()
num_folds <- 10
wae <- rep(0, num_folds)

for (t in 1:num_folds) {
  # *** THIS IS YOUR PREDICTION FUNCTION ***
  test_pred <- mypredict()
 
  # load fold file
  fold_file <- paste0('fold_', t, '.csv')
  new_train <- readr::read_csv(fold_file,
                               col_types = cols())

  # extract predictions matching up to the current fold
  scoring_tbl <- new_train %>%
      left_join(test_pred, by = c('Date', 'Store', 'Dept'))
  #readr::write_csv(scoring_tbl, paste0('scoring_', t, '.csv'))
  # compute WMAE
  actuals <- scoring_tbl$Weekly_Sales
  preds <- scoring_tbl$Weekly_Pred
  preds[is.na(preds)] <- 0
  weights <- if_else(scoring_tbl$IsHoliday, 5, 1)
  wae[t] <- sum(weights * abs(actuals - preds)) / sum(weights)
}

end.time = Sys.time()

```


#### Results


- 10-Fold Weighed Mean Absolute Error (WMAE)


```{r echo=FALSE, message=FALSE, warning=FALSE}
results = data.frame(Fold=seq(1,10), WeightedMeanError=wae)
results1 = data.frame(Mean_WAE=mean(wae))
knitr::kable(results) %>% kable_styling(bootstrap_options = "striped")

```


- Mean WMAE

```{r echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(results1) %>% kable_styling(bootstrap_options = "striped")

```


- Running time: System: MSI laptop, Intel i5, 2.0GHz, 8GB, Win 10

```{r echo=FALSE, message=FALSE, warning=FALSE}
results1 = data.frame(Run.Time=(end.time-start.time))
knitr::kable(results1) %>% kable_styling(bootstrap_options = "striped", font_size = 10)
```

#### Discussion

1. Log Transformation of Weekly Sales didn't yield desired result instead WMAE increased.

2. Since the given dataset didn't have many features as opposed to original dataset, it wouldn't be possible to predict negative weekly sales. So replaced negative weekly sales with zero to reduce WMAE.

3. Applied square root transformation yielded the desired with mean WMAE 1623 which was desirable.